---
title: "Comparing the calculation of MHWs in R vs Python"
author: "Robert W Schlegel"
date: "`r Sys.Date()`"
description: "This vignette compares the code/work flow one would use to calculate MHWs in both R and Python as well as the different outputs."
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.width = 4, fig.align = 'center',
                      warning = FALSE, 
                      message = FALSE, 
                      tidy = FALSE)
```

## Overview

The purpose of this vignette is to walk one through how to calculate MHWs in both the R and Python languages. It will use code from both languages throughout. The secondary purpose of this vignette is to show that the outputs are identical. The tertiary purpose of this vignette is to run benchmarks on the code to compare their speeds.

## Calculating events

First up, let's look at the two different ways in which one may calculate MHWs. First up, the R code, then Python.

### R code

#### Setup

The basic functionality for calculating marine heatwaves (MHWs) in R may be found in the __`heatwaveR`__ package that may currently be downloaded and installed with the following lines of code. Note that if one has already installed these packages they do not need to be installed again. To run one of the lines in the following code chunk will require that the hashtag first be removed.

```{r r-packages, eval=FALSE}
# install.packages("devtools")
## Development version from GitHub
# devtools::install_github("robwschlegel/heatwaveR") 
## Stable version from rOpenSci
## Currently not available
## Stable version from CRAN
# install.packages("heatwaveR") # Currently not available
```

With the necessary packages installed, we activate `heatwaveR` with the following line of code.

```{r r-init}
library(heatwaveR)
```

#### Default output

With everything downloaded and ready for us to use we may now calculate some events. The `heatwaveR` package has three built in time series (`sst_WA`, `sst_Med`, `sst_NW_Atl`) that we may use to more easily demonstrate how the code works. The general pipeline in `R` for calculating events is first to create a climatology from a chosen time series using `ts2clm()`, and then to feed that ooutput into `detect_event()`, as seen below.

```{r}
# First we create a default climatology as outlined in Hobday et al. (2016)
ts_clim <- ts2clm(data = sst_WA, climatologyPeriod = c("1983-01-01", "2012-12-31"))

# Then we feed that object into the second and final function
res <- detect_event(ts_clim)
```


### Python code

```{r python-setup, include = FALSE}
library(reticulate)
use_condaenv("ocean_py")
py_config()
```

#### Setup

To download and install the Python package for calculating MHWs one must run the following code.

#### Default calculations

## Comparisons

With some climatologies and events calculated with both languages we may now compare the results of the two.

```{r}

```

## Benchmarks

The final thing we want to look at in this vignette is the speed differences in calculating MHWs between the two languages.

```{r}

```

